---
title: 'Computer Workshop 5: Linear models with a categorical predictor'
author: "Roos & Pinard"
date: 'Last updated: `r Sys.Date()`'
output:
  word_document:
    toc: yes
    reference_docx: "../sbstemplate.docx"
---

# Workshop Structure

As always, we are using the workflow below to structure the workshops:

1. Formulate a research question.
2. Perform exploratory data analysis.
3. Identify any hidden assumptions.
4. Fit an appropriate model [Focus of today].
5. Diagnose the model and check assumptions.
6. Summarise the results.
7. Interpret and provide inferences.

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide", fig.show="show")
answer_sheet <- FALSE
library(ggplot2)
library(here)
```

A slight tweak today though, is that we're going to use the same dataset to run two models. The dataset is the same one you used in the first week of the course when we were looking at flat prices in Aberdeen.

The two models we're going to run are 1) a model that attempts to predict rent price as accurately as possible, and 2) a model that attempts to understand the causal effect that the square meters of a property has on on rent price.

The reason for doing this is that we want to make clear that models can have different purposes, depending on our needs. Want a fancy AI machine learning model? Sure, easy. But we can use the exact same machinery to also build a that understands causal effects.

We'll start with the predictive model, because they're much easier to do:

# 1. Formulate a Research Question

Fairly simple this time around. We just want to predict rent price. We don't care *why* a flat is more less expensive. We just want to be able to predict it accurately.

# 2. Perform Exploratory Data Analysis

As usual, carry out a quick data exploration looking for any mistakes. Keep in mind that you have a link to each property, so if you spot something that looks weird, you can resolve it yourself (a rare luxury!).

# 3. Identify Any Hidden Assumptions

Even though our aim here is pure prediction, we still want to think about hidden assumptions - we don't want to make predictions for flawed data. Spend a couple of minutes trying to identify any issues (or limitations, e.g. can you predict house prices in London from this data?).

# 4. Fit an appropriate model.

One of the things that makes predictive modelling so easy is that we can just throw all explanatory variables we have into the model with very little thought for what the underlying relationships would be. However, in the interest of keeping these instructions manageable, we'll specify that the model we want to run includes the explanatory variables:

* `sqmt` - the size of the property in square meters.

* `rooms` - the total number of rooms in the property (the sum of bedrooms and living rooms)

* `epc` - the energy performance of the property (categorical with levels A to G, note no A properties are present in the data)

* `tax` - the council tax band of the property based on assumed value of the property in 1991.

Notice that we're going to use `rooms` in the model, but you don't have this in your dataset. You'll need to create this new variable in your dataset. Try to figure out the code on your own but keep in mind that `rooms` is just `beds` added together with `living` (bedrooms and living rooms).

```{r, echo = FALSE}
abdn$rooms <- abdn$beds + abdn$living
```


The statistical notation for this model becomes a bit of a nightmare, simply because of how many levels there are in `epc` and `tax` but here's what it looks like:

$price_i \sim Normal(\mu_i, \sigma)$

$\mu_i = \beta_0 + \beta_1 \times sqmt_i + \beta_2 \times rooms_i + \beta_3 \times epcC_i + \beta_4 \times epcD_i + \beta_5 \times epcD_i + \beta_6 \times epcE_i + \beta_7 \times epcF_i + \beta_8 \times epcG_i + \beta_9 \times taxB_i + \beta_10 \times taxC_i + \beta_11 \times taxD_i + \beta_12 \times taxE_i + \beta_13 \times taxF_i + \beta_14 \times taxG_i$

where:

* $\beta_0$ is the intercept, representing a property with zero square meters, zero rooms, an EPC of B and a tax band of A.

* $\beta_1$ is the slope for each additional square meter

* $\beta_2$ is the slope for each additional room

* $\beta_{3,...,8}$ are the differences for each EPC relative to $\beta_0$

* $\beta_{9,...,14}$ are the differences for each council tax band relative to $\beta_0$

Like I said, a nightmare. You can see here why discretising (or creating groups) can be a problem in models. Instead of using a spectrum (or continuous value) for, e.g. energy efficiency, we're left with a model that now needs to estimate 14 parameters, rather than just four if we had continuous variables for EPC and council tax. As a general tip when you get to your honours projects and beyond, avoid "dichotomania" (or creating categoricals/dichotomies for data that's really a contuinuous spectrum) for your own sanity if nothing else.

Regardless, we can run this model fairly easily in `R`,

```{r}
rent_mod <- lm(price ~ sqmt + rooms + epc + tax, data = abdn)
```

Having run the model, let's see what the parameters $\beta_0$ and $\beta_1$ have been estimated as, using `summary()`:

```{r}
summary(rent_mod)
```

# 5. Diagnose the model and check assumptions.

In previous workshops, we've left assumption checks for the Friday session, but hopefully you've gained enough confidence now to do this relatively quickly.

Let's do a quick check now:

```{r}
par(mfrow = c(2,2))
plot(rent_mod)
par(mfrow = c(1,1))
```

## Question Set One

1. Do you see any assumptions that are likely broken in the Residuals vs Fitted and Scale-Location plots?

```{r}
# Yes, it looks like we have unequal variance of errors in both figures.
```

2. If so, what is the consequence of violating this assumptions?

```{r}
# Our confidence intervals are going to be artificially inflated, making our predictions look more uncertain than they should be.
```

3. Do you see any assumptions that are likely broken in the Q-Q Residuals plot?

```{r}
# Our residual errors are not perfectly Normally distributed but they're not that bad. We have three properties that are either more expensive, or cheaper, than we'd expect given the information we've provided to the model. If it were much more severe (i.e. dots all over the place), it can lead to inflated 95% CI.
```

4. Are there any observations that are having a disproportionate affect on the model?

```{r}
# Yes, a few. We can see these in the Residuals vs. Leverage plot. Particularly the properties on rows 114, 57 and 88 (note that 88 doesn't appear on the plot and is instead highlighted by a text output)
```

5. If there are any such observations, why might they be causing a problem?

```{r}
# We can check these properties by using the row number that the plot provides. To make it a bit easier, we can add in the predicted price
abdn$fit <- predict(rent_mod)
abdn[c(57, 88, 114),]

```




# 6. Summarise the results.

We'll come back to "5. Diagnose the model and check assumptions." on Friday. For today we'll go through how to take these results and convert them into a figure.

Just like last week, we'll use two options to plot our model results:

- Option 1: Use the `ggeffects` package to create the figure for you.

- Option 2: Create the prediction and figure yourself.

## `ggeffects`

Make sure you load `ggeffects` (and make sure it's installed if you didn't do so last week):

```{r}
library(ggeffects)
```

Then ask `ggeffects` to predict rabbit weight for each of our `mongoose` categories:

```{r}
preds <- ggpredict(rabbit_mod)
```

Now we just need to plot it:

```{r}
plot(preds)
```

Just like last week, we get a figure quickly and easily that shows our model predictions.

## Create the prediction and figure yourself.

We can also, broadly, repeat the process we used last week, with only a few minor tweaks to account for the different type of explanatory varibale (from continuous to categorical).

For starters, when we make our *fake* dataset, the values of `mongoose` need to only be `Present` or `Absent` (just like in our actual dataset). We don't need to create a range of values to predict over this time. To create our `mongoose` variable, we can either use `c("Absent", "Present")` or we can use the `unique()` function that you may have used in the very first Workshop. `unique()` will return all unique values within a column of data, so `unique(rabbit$mongoose)`, will also return `"Absent"` and `"Present"`. The benefit of using `unique()` is when we have categorical variables with many groups (or levels) and it becomes a real pain to write out all of the different groups.

Let's use `unique()` now to create our fake dataset to predict from:

```{r}
fake <- data.frame(
  mongoose = unique(rabbit$mongoose)
)
fake
```

Giving us a very simple dataset with just `mongoose` and two rows. One for `Absent` and one for `Present`.

Last week we used our `fake` dataset to only make a prediction for the "mean fit" (i.e. predicted boto length). This week we're also going to include uncertainty, specifically, we're going to also include 95% confidence intervals.

To do so, we need to add in some additional steps. For starters, because `predict()` is going to give us both the `fit` (i.e. predicted rabbit weight), but also `se.fit` (i.e. the standard error for the fit), we can't immediately add this to our `fake` dataset.

Instead, we'll just put the output from `predict()` into a temporary dataset, which we'll arbitrarily call `preds`:

```{r}
preds <- predict(rabbit_mod, newdata = fake, se.fit = TRUE)
```

The next part is the same as last week, just with an extra step. We can just go into the `preds` object and extract `$fit`, and include this in our `fake` dataset:

```{r}
fake$fit <- preds$fit
```

# Confidence intervals

This next part is new. We're going to use the standard error of the fit to calculate the 95% confidence intervals. If you need a reminder of what 95% CI are, then have a quick look at the lecture material.

The important thing to remember is that 95% CI take the standard error, multiplied by 1.96, and either *subtract* this from `fit` to get our *lower* 95% CI, or we *add* it to `fit` to get our *upper* 95% CI. Here's how we do that:

```{r}
fake$low95 <- preds$fit - 1.96 * preds$se.fit
fake$upp95 <- preds$fit + 1.96 * preds$se.fit
```

Now our `fake` dataset has all the values we need to do plot our results, using `geom_errorbar()` to add in the error bars to show our 95% CI:

```{r}
ggplot(fake) +
  geom_errorbar(aes(x = mongoose, ymin = low95, ymax = upp95)) +
  geom_point(aes(x = mongoose, y = fit)) +
  labs(x = "Java mongoose",
       y = "Predicted weight of rabbits (g)") +
  theme_minimal()
```

# The full workflow

Here's the entire workflow for the above analysis

```{r, eval = F}
# Deon Roos
# Analysis of Amami rabbit weight in relation to mongoose presence

# Load data
setwd("my/file/path")
rabbit <- read.table("rabbit.txt", header = T)

# EDA
summary(rabbit)
str(rabbit)
ggplot(rabbit) +
  geom_histogram(aes(x = weight))
ggplot(rabbit) + 
  geom_point(aes(x = mongoose, y = weight))
# Correcting 2 data entries identified in EDA and through communication with researchers
rabbit$mongoose[7] <- "Present"
rabbit$weight[rabbit$weight > 350] <- rabbit$weight[rabbit$weight > 350] - 100

# Concerns prior to modelling
# Unclear if mongoose on one island are actually absent or just not detected, or not reported
# Ideally this would be resolved before continuing with analysis but we are unable to

# Analysis
rabbit_mod <- lm(weight ~ mongoose, data = rabbit)

# Diagnostics
# Covered in next workshop

# Inference
summary(rabbit_mod)

fake <- data.frame(mongoose = unique(rabbit$mongoose))
preds <- predict(rabbit_mod, newdata = fake, se.fit = TRUE)
fake$fit <- preds$fit
fake$low95 <- preds$fit - 1.96 * preds$se.fit
fake$upp95 <- preds$fit + 1.96 * preds$se.fit
ggplot(fake) +
  geom_errorbar(aes(x = mongoose, ymin = low95, ymax = upp95)) +
  geom_point(aes(x = mongoose, y = fit)) +
  labs(x = "Java mongoose",
       y = "Predicted weight of rabbits (g)") +
  theme_minimal()
```

## Question Set Five

1. Once you've completed this analysis (not including model diagnostics, which we'll cover on Friday), in groups, discuss whether you think these results should be shared with the local government as evidence to help them decide if mongoose eradication should continue or not.

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# As said above, personally I have no idea what to make of this model given that the assumption of validity seems questionable (related to our uncertainty of mongoose being present or not in Tokunoshima). 

# Making it even harder, the apparent effect of mongoose is tiny, just two gramms! - small enough that it could be something as silly as using two different sets of scales on both islands.

# The thing that would keep me up at night would be if anyone were to use our model to make any decisions that could have real world consequences, such as showing these results to the local government. Are we really confident enough in our model to have that rest on our conscious? I don't think I am.

# This last part is probably the hardest part of "statistics" and it has nothing to do with actual statistics; the equations, the distributions, etc, none of that matters now. It's all down to asking if the data is appropriate for answering our question.
```

# A new challenger

Once you're finished with the Amami rabbit dataset, move on to the `lemur.txt` dataset on MyAberdeen. This dataset comes from researchers who are interested in understanding how activity levels (i.e. how many hours a group of lemurs are active for in a day), varies across four habitat types.

The data contains:

- `activity` - the number of hours lemurs were active for (measured to two decimal places)
- `habitat` - one of four habitat types; Rainforest, Dry Forest, Spiny Forest, and Mangrove.

Go through the workflow you've just refined for fitting a linear model with a categorical explanatory variable and see which habitat type lemurs are most active in.

```{r, include = F}
# For coordinators -
# Code below simulates and exports the data used in the workshop
# number of offspring as function of temperature

set.seed(2024)
n <- 120
habitat_types <- factor(rep(c("Rainforest", "Dry Forest", "Spiny Forest", "Mangrove"), each = n / 4))

# lazy way to simulate the data
activity_levels <- round(c(
  rnorm(n/4, mean = 10, sd = 1),   # Rainforest group
  rnorm(n/4, mean = 8, sd = 1),    # Dry Forest group
  rnorm(n/4, mean = 6, sd = 1),    # Spiny Forest group
  rnorm(n/4, mean = 7, sd = 1)     # Mangrove group
), digits = 2)

# Create the data frame
lemur <- data.frame(habitat = habitat_types, activity = activity_levels)
lemur$habitat <- as.character(lemur$habitat)
lemur$habitat[23] <- "Rain forest"

write.table(lemur, here("Workshops/Workshop 5 - LM with categorical predictors/Data", "lemurs.txt"), row.names = FALSE)
```

# Assumption table


| **Assumption**                   | **Description**                                                                                             | **Example of Violation**                                                                                                                                       |
|---------------------------------|-------------------------------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| 1. **Validity**                     | The data is able to answer the research question effectively.                      | Using IQ to measure overall intelligence (what does IQ even measure?).                      |
| 2. **Representativeness**           | The data reflects the population or group being studied.                              | Using survey data from a rural area to draw conclusions about an urban population. |
| 3. **Additivity**                   | Effects of predictors are additive and there are no interactions between them.         | Assuming that study hours and sleep hours independently affect exam scores without considering their combined effect (imagine a Venn diagram with study and sleep overlapping). |
| 4. **Linearity**                   | The relationship between predictors and the outcome is linear.                                   | Assuming plant growth will continue to increase indefinitely without ever levelling out.          |
| 5. **Independence of Errors**       | Observations are independent of each other. (Related to residual errors.)                                                       | If modelling height, the data come from the same group of children over time.               |
| 6. **Equal Variance of Errors**     | Residuals have constant variance across all levels of predictors ("homoscedasticity"). (Related to residual errors.)  | If modelling income, variability in salaries is not the same across all levels of a company’s hierarchy.              |
| 7. **Normality of Errors**          | Residuals should be normally distributed.  (Related to residual errors.)                         | If modelling income, the presence of any extremely wealthy individuals will be hard for the model to understand and explain.     