---
title: 'Computer Workshop 1: Getting back into R'
author: "Roos & Pinard"
date: 'Last updated: `r Sys.Date()`'
output:
  word_document:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    code_folding: show
    theme: readable
---

# Workshop Structures

The workshops on this course will have you learn an analytical workflow, focussing mainly on analysing data using linear models (and eventually generalised linear models). That is, you are going to learn how to go from getting a dataset to extracting evidence to answer your research question. Importantly though, this workflow is not unique to linear models. You would (or at least *should*) use this same general workflow, regardless the scenario, the analytical method, or whichever industry you may be faced with.

The general workflow we'll be using can be broken down into:

1. Think of a research question
2. *Carry out exploratory data analysis*
3. Fit an appropriate model
4. Diagnose the model and check assumptions (that we are able to)
5. Summarise the results
6. Interpret and provide inference

For these workshops we tend to skip Step 1 in the workflow for convenience sake, but in your own work Step 1 is key as every subsequent step in the workflow is dependent on your question.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # uncomment if producing answer sheet
#knitr::opts_chunk$set(echo = TRUE, results = "hide", fig.show="hide") # uncomment if producing workshop sheet
answer_sheet <- TRUE # Put this to TRUE to generate the workshop sheet with answers, FALSE to generate the workshop sheet without answers
library(ggplot2)
library(here)
```

# Learning objectives

1. (Re)familiarise yourself with the basics of `R`;
2. Use `R` to do simple descriptive summaries of data;
3. Consider the results of the EDA (Exploratory Data Analysis) to identify any issues;

All three of these objectives link to useful skills for doing initial and quick explorations of a new data set. All of these skills will be used through out the course.

```{r datasets, echo = FALSE, message = FALSE}
# For coordinators -
# Code below simulates and exports the data used in the workshop
# alligator bite strength
# Mean is 2000 lbs (from google)
# Mean length is 300 cm to 460 cm
# Simulated as:
# bite ~ Normal(mu, 60)
# mu = b0 + b1 * length
set.seed(2024)
n <- 78
ali <- data.frame(
  length = rnorm(n, mean = 350, sd = 60)
)
b0 <- 0
b1 <- 5.7
mu <- b0 + b1 * ali$length
ali$bite = rnorm(n, mean = mu, sd = 60)
# Creating a data entry mistake caused by a "missing" decimal point
# Observation 20 is problem sample
# ggplot(ali) +
#   geom_histogram(aes(x = length), colour = "white")
# ggplot(ali) +
#   geom_histogram(aes(x = bite), colour = "white")
ali$length <- round(ali$length, digits = 1)
ali$bite <- round(ali$bite, digits = 2)

mistake <- 20
ali$bite[mistake] <- ali$bite[mistake] * 100
ali$length[mistake] <- ali$length[mistake] * 10

write.table(ali, "H:/003 - Teaching/006-BI3010/Workshops/Workshop 1 - Hidden assumptions/Data/alligator_bite.txt", row.names = FALSE)
```

# Instructions
In this workshop, we're going to be loading in various datasets, carrying out an initial exploration and do "Quality Assurance". When doing real-world analysis, this is an incredibly important step because mistakes are easily made when doing data entry; Imagine entering in 500 near identical rows into an Excel document - it's hard to stay super focused the entire time so mistakes happen regularly. Those mistakes can be a serious problem in analysis, hence why we want to check if this has happened.

The good news for us is that we don't need any crazy, complicated stats or `R` coding to do some initial screening. We can just load the data into `R` and do some quick checks. For this workshop, this will also help to reintroduce you to `R` and get back into the swing of things.

We would like you to use RStudio.  If you have not used this before, please use the instructions on MyAberdeen called "Getting Started with R and RStudio".

## Getting started

### Set working directory
If you haven’t already created a folder for the course on your computer or your university drive, do so now.  You might create a folder for the course (e.g., 'BI3010') with subfolders within that for each of the workshops (e.g., workshop1).

To set the working directory to the folder you want to use for saving your work, you have two options:

Option 1:

Find the file path for your newly created folder (e.g. `BI3010` > `workshop1`). An easy way to get this is by using File Explorer in Windows, go to your folder, then highlting the address bar. It will probably look something like: 

`H:\BI3010\workshop1`

In `R` backslashes are used for specific purposes, so we want to change the back slashes for forward slashes, so our file path looks like:

`H:/BI3010/workshop1`

We can now write some `R` code to tell `R` that this is the folder that we will be using in the current session, by setting the Working Directory using the function `setwd()`. The `setwd()` function just needs a single piece of information (technically called an "argument"), which is the file path, but because the file path is a "string" (i.e. a bunch of letters, characters and numbers) we need to ensure that we put the file path in quotation marks:

`setwd("H:/BI3010/workshop1")`

The benefit of doing it this way is that we can include this code in our `R` script, so that we don't have to go through this pain every time we come back to work on this.

Option 2:

An easy option is to use the drop down menu in RStudio: session > set working directory > choose directory

This is easier to do the first time, but becomes a pain when you come back to the project, as you'll need to reset the working directory (and maybe make a mistake, causing more pain and frustration).

We'd advise using Option 1. A little more of a fuss the first time but then you don't need to think about it again.

We can check if the working directory has been set correctly by asking `R` what the current directory is:

```{r, eval = F}
getwd() #to check which directory RStudio is using
```

### Download the file

**Download the datafile from MyAberdeen to your working directory.**

You will find a file called `alligator_bite.txt` in the module for Week 1, in the folder for Workshop 1. Using the three dots to the right of the content area that has the file name, download the file and save it to your working directory

### Load file into RStudio

Now that we've told `R` which folder contains our data, we can load it in. Before we do so, make note of what type of file the data is saved as. The alligator bite strength data is saved as a text file, which we know because of the `.txt` extension. For data saved in this format, we use the `read.table()` function, where the arguments are. To find out which arguments can be provided to a function you can type `?read.table`. If we do so, we can see there are lots of arguments we could use, but the crucial ones are `file` (which is the name of the file we want to load), and another called `header` which is logical (meaning either `TRUE` or `FALSE`) which tells `R` if the first row of our `alligator_bite.txt` file are the column names, which they are.

So to load our data, we use the following code:

```{r, eval = FALSE}
ali <- read.table("alligator_bite.txt", header = TRUE) 
```

If you get an error, some common problems are as follows: you may have set the wrong working directory; you may have misspelled one of the words (including using a capital letter instead of lower case); your file is missing a column heading; the file is not in the correct format (e.g, the file is saved as a `.csv` and not a `.txt`); the quotation marks are in the in wrong location; or you may have missed the final bracket “)” in the `read.table()` function.

To confirm that the file has been read properly, double click the file name in the Environment panel (normally in the upper right of RStudio).

If you want to visually look at the data now that it's in `R`, you have a couple of options:

The following will show you the first six rows of your dataset (you can change how many rows by using the argument `n`, e.g. `head(ali, n = 10)`):

```{r}
head(ali)
```

Or you can open the entire dataset in a new tab by doing:

`View(ali)`

The output of `View()` has the benefit of being visually similar to Excel spreadsheets, with many of the same features such as allowing the user to sort and filter by columns.

Or you can output all rows into the console, simply by doing:

`ali`

We wouldn't recommend this last option as it's not a particularly useful one, given that it floods your console, making it hard to see anything of use.

### Question Set One

Question 1: How many variables do we have in this dataset?

Question 2: What are the variable names?

Question 3: Are the names captilised?

These questions may seem trivial, but honestly, they're the first questions we ask ourselves when we load in a dataset to `R` for the first time.

## Getting to know the data

Here's what you know about the data. The data contains `nrow(ali)` observations (sometimes refered to as samples) of `nrow(ali)` alligator bite strength, measured in pounds per square inch. All `nrow(ali)` alligators were measured in Parker Island Gator Farm, Florida, USA. For each bite strength, the length of the alligator, from nose to tail, was measured and is provided in centimeters (to one decimal place).

Before we move ahead with the data, let's take a second to become familiar with alligator bite strength. Using your search engine of preference, do a quick search to find out the bite strength of alligators. Specifically

When we do exploratory data analysis, we're specifically looking for things that look *weird*. In order to know if something is *weird*, we need to know what "normal" is; hence why it's a good idea to have some sort of notion of what a "normal" alligator's bite strength may be.

Write down the average bite strength of an alligator, measured in pounds per square inch and the average length of an alligator in centimeters.

With that done, let's move onto understanding our data. For starters, let's check if the data is sensibly structured. In `R` this means that numbers are numbers, words are characters, and so on. Basically, we want to make sure there weren't any bite measurements like `2174.94BIGBITE`, which will mess us up by meaning that `R` thinks that all `bite` values are characters and not numbers.

To do this quickly, we can use the `str()` function (short for "structure"):

```{r}
str(ali)
```
We see that `ali` is a `data.frame` (great - it should be!), with 78 observations (or samples) and two variables. The `$` is used to extract specific variables of a dataframe, so `ali$bite` will give you just the `bite` variable, hence why the second and third line are proceeded by `$`. From this, we can see that both the `length` and `bite` variables are being treated as numbers (the `num` part), and we then get a handful of example values for both `length` and `bite`.

So far, we're not seeing anything unusual which is great - it means we have not identified any problems that we need to fix.

If we so wanted to, we could extract the `length` variable from the dataset, by doing:

```{r}
ali$length
```

How would you extract the `bite` variable?

Let's use these extracted variables to calculate some summary statistics, starting with the mean (i.e. average). To do so, we use the `mean()` function. To find the average length of our alligators, we do:

```{r}
mean(ali$length)
```

We can see that the average crocodile in Parker Island Gator Farm is `r round(mean(ali$length), digits = )` cm long. 

Compare this value with the mean length of alligators you found from a quick internet search. A little bit longer than we might have expected but maybe it's something to do with these alligators coming from a farm, as opposed to wild ones. Maybe they're given lots of food to help them grow? Who knows, but we wouldn't expect our numbers to match exactly with whatever number we find online, right?

Do the same thing for `bite`; Calculate the average bite strength of our alligators, and let's see if there's anything weird going on there.

```{r}
mean(ali$bite)
```

Assuming you also found that alligator bite at roughly 2,000 pounds per square inch, our alligators seem to be biting *on average* with an extra 1,500 pounds per square inch. That seems like *a lot* more. 

This is something that should make us sit up and start paying a little more attention. We don't know there is a problem, but we have reason to believe there *might* be one. It's now up to us to prove that there *isn't* a problem.

There are lots of descriptive techniques we can use to check for odd things in our data, but a convenient "short cut" option is to use the `summary()` function. This function will provide a generic summary of our entire dataset if we supply `ali` as the sole argument. Do so now and let's see if we can track down what's going on in this data.

```{r}
summary(ali)
```

Notice anything weird in the above summary?

### Question Set Two

Question 1. How small is the smallest alligator in the farm for which we have data?

Question 2. What is the biggest?

Question 3. What are the minimum and maximum values for bite strength?

## Data detective

Hopefully you've identified that there is at least one questionable value in our dataset. What do we do now? Throw out the entire dataset? (No... The answer is "No you do not throw out the dataset" - at least for now). Let's try and figure out what's going on here, starting with finding the problem observations.

How do we do that?

Let's start with considering what we think a weird observation may look like. We can use the `summary(ali)` output to help us here. Starting with `length`, we know we have alligator that was apparently `r max(ali$length)` cm long (`r round(max(ali$length)/100, digits = 0)` m - that's longer than a bus!). How do we find that value in our dataset to see if there was anything else weird about it.

This is where we can use the square brackets `[]`. In `R`, square brackets allow us to extract specific observations from, in this case, a dataframe. To use them, first think how you would do this by hand:

1. We'd start by looking at the `ali` dataset
2. Within the `ali` dataset, find any rows where alligators are `r max(ali$length)` cm long
3. For any rows where it's true that the length is equal to `r max(ali$length)` cm, then show me all of the information related to that alligator

How does that translate into `R`? Almost exactly - we just write those three steps using code.

1. We'd start by looking at the `ali` dataset

```{r}
ali
```


2. Within the `ali` dataset, find the rows where alligators are `r max(ali$length)` cm long. 

This is where we can make use of the square brackets as well as conditional statements. Conditional statements are where we can say "when this condition is met, do *something*". In our case, our condition is an alligator that's `r max(ali$length)` cm long. In `R` we can write this condition as (note that we use double equal signs to say "equal to"  in `R`):

```{r}
ali$length == 123632.0
```

The output is a list of `r nrow(ali)` values of `TRUE` and `FALSE` values, one for each row of our `ali` dataset. Any `TRUE` values indicate an alligator that is equal to `r max(ali$length)` cm long, and `FALSE` values are when that is not the case (in `R` we write "not equal to" as `!=`).

3. For any rows where it's true that the length is equal to `r max(ali$length)` cm, then show me all of the information related to that alligator

Here is where we can use the square brackets. Square brackets allow us to extract specific rows, columns or just individual values from datasets. When using square brackets keep in mind that row number goes before the comma and column number goes after (e.g. `ali[row,column]`). If we don't specify a value for either `row` or `column`, `R` will give us all of them. For example, if we want the value from the 30th row and second column, we would do:

```{r}
ali[30,2]
```
We can use these square brackets, in combination with the conditional statement, to extract the entire row for the alligator that's `r max(ali$length)` cm long. What happens in this case is the `TRUE` and `FALSE` values act as a way to tell `R` which rows we want to see. If it's `TRUE`, then we want to see it. If it's `FALSE`, then don't show us. We can do this by writing:

```{r}
ali[ali$length == 2293,]
```

What do you make of this observation? We're expecting `length` to be about 200 to 400 centimeters (2 to 4 m), but this alligator is apparently 10 times bigger than that. And `bite` is about 100 times bigger.

Why might that be the case? We'll let's see if we can find a "normal" observation. Let's assume the first observation is the normal one. Use the square brackets to extract the first row of data from the `ali` dataset.

```{r}
ali[1,]
```

Well they're clearly different. There are only two possible explanations for this. Either there are giant alligators among us. Or, there has been a data entry mistake.

Using the above, and whichever other methods you can think of, try to figure out what the potential reason is for this observation of an apparently weird, super big alligator.

### Question Set Three

Question 1. What may have occurred to have caused this apparent data entry mistake?

```{r, include=answer_sheet, class.source="bg-success"}
# This "giant" alligator is the result of a data entry mistake whereby the decimal points were missing for bite and length. The values should be 229.3 cm long and	1236.32	pounds per square inch. If you look at the raw data, you'll see it has no decimal points.
```

## A more refined detective

The way we identified the above data entry mistake is really useful, but what if we wanted to be a slightly more refined programmer in checking that observation? Basically, what if we didn't want to write out `r max(ali$length)` every single time?

What we can do instead, is a functions that says what the maximum alligator length was. The function to do just that is called `max()`, and it works the same way as `mean()`, but instead it will tell you the maximum rather than the mean:

```{r}
max(ali$length)
```

We can use that when extracting the problematic variable by doing:

```{r}
ali[ali$length == max(ali$length),]
```

Which you can read as saying: "Within the `ali` dataset, extract the rows where the length of an alligator is equal to the maximum length of any alligator".

We can also adapt the conditional statement. For instance, if we wanted to extract all rows with alligators that were *greater* than the average, we could do:

```{r}
ali[ali$length > mean(ali$length),]
```

We now get data from more than one alligator as, unsurprisingly, more than one alligator meets our condition of being longer than average. We can also see the "data entry mistake alligator" using this method.

Let's keep refining this method by using a different function; `quantile()`. Go back to the `summary(ali)` output and have a look at the quantiles (`1st Qu.` and `3rd Qu.`) that it reports. These represent the values above and below which contain 25% and 75% of the data. We can get these values by doing:

```{r}
quantile(ali$length, probs = c(0.25,0.75))
```

The `probs` argument is where we can specify what probabilities we would like quantiles for. In the above example, we've asked for the 25% (`0.25`) and 75% (`0.75`) quantiles. But because we've asked for two values, we need to "wrap" these in another function - the `c()` function. `c()` combines, or concatenates, more than one value together, hence how we were able to get to values of our the `quantile()` function. If we only wanted a specific quantile, we could write `quantile(ali$length, probs = 0.75)`. Because we only want one value, we don't need to use `c()`.

### Question Set Four

Using `quantile()` extract all rows in the `ali` dataset where `bite` values are above the 3rd quantile.

Repeat the above, but for `bite` values *less than or equal to* the 33% quantile. If you're unsure how to specify *less than or equal to* have a quick search online or use your preferred LLM.

Using the tools from above, can you identify any other possible data entry mistakes?

## Be the creator










