---
title: 'Computer Workshop 1: Getting back into R'
author: "Roos & Pinard"
date: 'Last updated: `r Sys.Date()`'
output:
  word_document:
    toc: yes
    reference_docx: "../sbstemplate.docx"
---

# Workshop Structure

The workshops on this course will have you learn an analytical workflow, focussing mainly on analysing data using linear models (and eventually generalised linear models). That is, you're going to learn how to go from getting a dataset to extracting evidence to answer your research question. Importantly though, this workflow is not unique to linear models (or GLMs). You would (or at least *should*) use this same general workflow, regardless the dataset, the scenario, or the analytical method you may find yourself using.

The general workflow we'll be using can be broken down into:

1. Think of a research question
2. Carry out exploratory data analysis [Focus of today]
3. Identify any hidden assumptions
4. Fit an appropriate model
5. Diagnose the model and check assumptions
6. Summarise the results
7. Interpret and provide inference

For the BI3010 workshops we are generally going to skip Step 1 in the workflow for convenience sake, but in your own work Step 1 is key as every subsequent step in the workflow is dependent on your question. In your own work (e.g. your Honour's project) it's important to take the time the think carefully about what the question you want to answer is, how best to design an experiment or data collection protocol to get data suitable for answering it, which statistical method would be suitable for analysing the data, and what caveats and limitations come with all of these. Basically, don't take the omission of step 1 in this course as evidence that it's not important. It is.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
answer_sheet <- TRUE
library(ggplot2)
library(here)
```

# Learning objectives

1. (Re)familiarise yourself with the basics of `R`;
2. Use `R` to do simple descriptive summaries of data and data processing;
3. Use the results of the EDA (Exploratory Data Analysis) to identify any issues;
4. Begin thinking about what assumptions we make with a specific dataset even before we begin using statistics

All of these objectives link to useful skills for doing initial, quick explorations and screening of data sets. Because it's so important to ensure that the data we subsequently use in analysis is suitable for answering our question, the importance of such EDA cannot be overstated.

These skills will be used through out the course.

```{r datasets, echo = FALSE, message = FALSE}
# For coordinators -
# Code below simulates and exports the data used in the workshop
# alligator bite strength
# Mean is 2000 lbs (from google)
# Mean length is 300 cm to 460 cm
# Simulated as:
# bite ~ Normal(mu, 60)
# mu = b0 + b1 * length
set.seed(2024)
n <- 78
ali <- data.frame(
  length = rnorm(n, mean = 350, sd = 60)
)
b0 <- 0
b1 <- 5.7
mu <- b0 + b1 * ali$length
ali$bite = rnorm(n, mean = mu, sd = 60)

# Plots to check sim data (without mistake below)
# ggplot(ali) +
#   geom_histogram(aes(x = length), colour = "white")
# ggplot(ali) +
#   geom_histogram(aes(x = bite), colour = "white")

# Rounding
ali$length <- round(ali$length, digits = 1)
ali$bite <- round(ali$bite, digits = 2)

# Creating a data entry mistake caused by a "missing" decimal point
# Observation 20 is problem sample
mistake <- 20
ali$bite[mistake] <- ali$bite[mistake] * 100
ali$length[mistake] <- ali$length[mistake] * 10

# Exporting so data used here is same as the one students have
# Should always be the same unless values above are changed (including seed)
write.table(ali, here("Workshops/Workshop 1 - Hidden assumptions/Data", "alligator_bite.txt"), row.names = FALSE)
```

# Coding style

All instructions for the workshops were made in `R` using something called `RMarkdown`, which were then exported as Word documents. Within these instructions, code will either being `inline` (meaning the code appears in the normal text line, `like this`) or else in "chunks". When in chunks they will have the following colour coding:

```{r, eval = F}
# Comments are not read as code.
# They are notes for ourselves and others.
my_result <- a_function(argument1 = "character1", 
                        argument2 = variableA,
                        argument3 = 1234,
                        argument4 = c(TRUE, FALSE),
                        argument5 = my_data)
```

* The light purple text preceded by "#" are comments. These will not be treated by `R` as code, so we are free to write whatever we want. See below for guidance on comment writing.

* The orange code (e.g. `a_function()`) denotes a function. Functions are blocks of re-usable code, often written by someone else, to perform a specific task, for example `mean()`. Underlying each function is a collection of "source" code to do the the specific task (if you want to see the source code for `mean()` write `getAnywhere(mean.default())` in your console). Functions are always followed by an open parenthesis "(" and ended with a closed parenthesis ")".

* The pale blue code (e.g. `argument1 =`) is used to indicate arguments - data, parameters or choices that the function requires, or allows, to perform it's purpose. You can think of arguments as the information required, or permitted, for the function to do it's task. Multiple arguments are separated using commas.

* Green code, wrapped in quotation marks (e.g. `"character1"`), indicate strings or characters which are types of data used to represent text.

* Black code (e.g. `variableA`, `my_result` and `my_data`) indicate objects in the R environment. Generally, these are stored single values, columns of data or even entire datasets.

* Finally, the maroon code (e.g. `TRUE` and `FALSE`) are so-called "Boolean" values. These are some of the simplest forms of data and are used to indicate, unsurprisingly, whether something is true or false (commonly used in functions to allow the user to say "Yes" or "No").

## Commenting guide 

It's good practice to make comments to yourself (and others) while writing code. Comments generally act as notes that help explain your general intent for a section of code, which in three weeks time, can be an absolute life saver. They're also super useful for anyone else reading your code (and your commenting may be evaluated for some jobs).

There are many different opinions as to how to write good comments (see [here](https://statmodeling.stat.columbia.edu/2024/02/07/when-all-else-fails-add-a-code-comment/) for a particularly stern view), but a general principle I subscribe to is; aim to very concisely describe your general purpose for the next few lines of code. What are *you* trying to do, and not, what does *each line of code* do.

For example, I would write:

```{r, eval = F}
# Get R set up
library(ggplot2)  # For plots
setwd("my/file/path")
my_data <- read.csv("data/sales_data.csv")

# Plot the data
ggplot(my_data, aes(x = country, y = total_sales)) +
  geom_bar()
```

And not:

```{r, eval = F}
# Load the ggplot2 package into the R environment so we can create plots later
# What happened to ggplot1?
library(ggplot2)

# Use the read.csv function to read the sales_data.csv file from the data directory
# and store the resulting data frame in a data frame called 'my_data'
# Remember that it's called my_data and not data. Calling it data made things break :(
my_data <- read.csv("data/sales_data.csv")
# my_data <- read.table("data/sales_data.csv") # DONT USE THIS IT DOESNT WORK
# mi_data <- read_csv("data/sales_data.csv") # ALSO DIDN'T WORK - DUNNO WHY

# Use ggplot to create a bar plot of the total sales by country for 2024.
# The x-axis represents the categories, and the y-axis represents the total sales.
ggplot(my_data, aes(x = country, y = total_sales)) + # using ggplot, take my_data, then plot country on the x and sales on the y
  geom_bar() # then have the data visualised as bar, using the structure described in the previous line of code
# When I see comments like this IRL, I die a little bit inside. Keep it clean and concise.
```

The latter is so overly descriptive that it becomes as much of a chore to read the comments as the actual code itself. In fact, I think the code does a better job quickly and efficiently explaining its purpose than these comments do! Another way to think of how to best write comments is; Think of comments as post-it notes. You want them to be short, and to act as a signal for sections that you want to dive deeper into (e.g. for debugging).

## On using LLMs

LLMs (Large Language Models), or "AI", are incredibly useful tools for programming. Whenever I am coding, I will normally have a LLM open and ready for any requests I might have. Personally, I find them particularly useful for 1) doing boring or fiddly coding, 2) automating tasks 3) to solve one-off problems and 4) to simplify and tidy my code (see [here](https://nicholas.carlini.com/writing/2024/how-i-use-ai.html) for a pragmatic discussion on the use of LLMs in coding). But using them comes with the requirement that you have to know enough about coding to know when the answer they're giving you is useful or when they're... not so useful. We're not quite at the stage where "you don't have to know how to write code to write code", simply because a lot of the time the code LLMs produce simply doesn't work. As in, it's just nonsense. Especially as the complexity of the code increases beyond "basic". But the benefit for you guys is that learning curve of coding has definitely flattened.

Using LLMs effectively, however, is somewhat a skill in itself, in the same way as knowing what code or error to Google before LLMs came about. As a tip, make your prompt as detailed as possible. Describe in simple terms what you want to achieve, include the relevant code you tried to use, include any errors, and if possible, what you think the problem might be. It's also worth telling the LLM that you are working in `R` (so it doesn't give you `python` code), and to mention any packages (more on these on Wednesday) you want to use, or, alternatively, if you want to use base version `R`.

From bitter experience, I'd strongly recommend having a separate script open to test the code provided by an LLM before replacing your own code. I've been bitten too many times with flawed and broken LLM code that destroys entire scripts because I immediately replaced my original code before testing the LLM's version. Test first, then combine.

# Workshop Instructions

In this workshop, we're going to be loading in various datasets, starting with alligator bite strength, and carrying out so-called Exploratory Data Analysis (EDA) and performing some simple Data Quality Assurance. Broadly, our aim in such cases is to try and find *weird* things in the data which imply mistakes, such as data entry mistakes. When doing real-world analysis, this is an incredibly important step of analysis because mistakes are so easily made, especially when doing data entry. Imagine entering in 5,000 near identical observations into an Excel document. People get bored, start thinking about other things, get mixed up over which observation they are meant to be adding in, amongst many other distractions. Basically, it's hard to stay super focused doing mundane work for long periods of time. It'd be silly to expect or assume these mistakes would never happen, so we check for them. The problem is that if don't spot them, any mistakes can be hugely influential when it comes time to analyse the data.

The good news for us is that we don't need any crazy, complicated stats to check for this, and the `R` coding is *relatively* straightforward. The first step is just loading the data into `R`.

We would like you to use RStudio, an Integrated Development Environment. Basically, a program that makes interacting with `R` (the programming language) a much more pleasant experience. If you have not used RStudio before, please use the instructions on MyAberdeen called "Getting Started with R and RStudio".

## Getting started

### Set working directory

If you haven’t already created a folder for the course on your computer or your university drive, do so now. For example, you might create a folder in your `H` drive for the course, with subfolders for each of the workshops (e.g., `workshop1`), such that the file path to your working directory would be: `H:\BI3010\workshop1`.

Open RStudio and create a new R script (`File` > `New File` > `R Script`, or use the hotkeys `Ctrl + Shift + N`). Once open, you can now write code in the script. When you want to run that code, either press the Run button (top right of the main panel) or click anywhere in a line of code use the hotkeys `Ctrl + Enter`. 

With your script now open, to set the working directory to the folder you want to use for saving your work, you have two options:

**Option 1:**

Find the file path for your newly created folder (e.g. `H:` > `BI3010` > `workshop1`). If you're not sure, an easy way to get this is by using File Explorer in Windows, go to your `workshop1` folder, then select the address (or file path) from the navigation bar at the top (the equivalent of the URL bar in web browsers). It will probably look something like: 

`H:\BI3010\workshop1`

In `R` backslashes are used for specific purposes, so we want to change the back slashes for forward slashes, so getting our file path ready for `R`, it should look like:

`H:/BI3010/workshop1`

We can now write some `R` code to tell `R` that this is the folder that we will be using in the current session, by setting the Working Directory using the function `setwd()`. The `setwd()` function just needs a single piece of information (technically called an "argument"), which is the file path, but because the file path is a "string" (i.e. a bunch of letters, characters and numbers) we need to ensure that we put the file path in quotation marks:

```{r, eval = FALSE}
setwd("H:/BI3010/workshop1")
```

The benefit of doing it this way is that because we can include this code in our `R` script, we don't have to go through this pain of setting the working directory every time we come back to work on this script (it also means we can copy and paste this code when we get to subsequent workshops, just changing workshop1 with e.g. workshop2).

**Option 2:**

An easy option is to use the drop down menu in RStudio: `Session` > `Set Working Directory` > `Choose Directory...`

Option 2 is easier if you're just doing a very quick project (i.e. an hour or two), but becomes a pain when you come back to the project, as you'll need to keep setting the working directory each time (and maybe make a mistake, causing more pain and frustration).

We'd advise using Option 1. A little more of a fuss the first time but then you don't need to think about it again.

We can check if the working directory has been set correctly by asking `R` what the current directory is:

```{r, eval = F}
getwd() #to check which directory RStudio is using
```

### Download the file

*You can download the data from MyAberdeen. Make sure to save the file to your working directory.*

On MyAberdeen in the Tuesday - Workshop folder within Week 1, you'll find a file called `alligator_bite.txt`. Either click on the file to download it or, if that isn't working, use the three dots to the right of the content area that has the file name to download it.

Once downloaded, make sure to move it to your `H:/BI3010/workshop1` folder (or whatever you chose to call your folders).

### Load file into RStudio

Now that we've told `R` which folder contains our data (by using `setwd()`), we can load the data into `R`. Before we do so, make note of what type of file the data is saved as. The alligator bite strength data is saved as a text file, which we know because of the `.txt` extension. For data saved in this format, we use the `read.table()` function, where the arguments are predominantly what the file is called, and if the file contains headers. 

To find out which arguments can be provided to a function you can type `?read.table` (or `?any_function_here`). If we do so, we can see there are lots of arguments we could use, but the crucial ones are `file` (which is the name of the file we want to load), and another called `header` which is logical (meaning either `TRUE` or `FALSE`) which tells `R` if the first row of our `alligator_bite.txt` file are the column names, which they are.

So to load our data, we use the following code:

```{r, eval = FALSE}
ali <- read.table("alligator_bite.txt", header = TRUE) 
```

Note that this is a single line of code on a single line of script. But there's nothing to stop us from spacing this out to make it easier to read, if we so wish:

```{r, eval = FALSE}
ali <- read.table("alligator_bite.txt", 
                  header = TRUE) 
```

There are a few things to make sure of when doing this, but just experiment. The great thing about working in `R` is that nothing is permanent. If you break something, you can just run your code again. The only thing that is "permanent" is your script, hence why it's important to save often (and also to write comments to yourself). Added something new? Save and write a comment saying what it does. Want to remove something but not sure if it'll break your code? Put a # at the start of the line to "comment it out".

When loading your data in, if you get an error, some common problems are: 

* You may have set the wrong working directory; 

* You may have misspelled one of the words (including using a capital letter instead of lower case); 

* Your file may be missing a column heading; 

* The file is not in the correct format (e.g, the file saved as a `.csv` and not a `.txt`); 

* The quotation marks are in the in wrong location; 

* You may have missed the final bracket “)” in the `read.table()` function.

To confirm that the file has been read properly, double click the file name in the Environment panel (normally in the upper right of RStudio).

If you want to visually look at the data now that it's in `R`, you have a couple of options:

The following will show you the first six rows of your dataset (you can change how many rows by using the argument `n`, e.g. `head(ali, n = 10)`):

```{r}
head(ali)
```

Or you can open the entire dataset in a new tab by doing:

`View(ali)`

The output of `View()` has the benefit of being visually similar to Excel spreadsheets, with many of the same features such as allowing the user to sort and filter by columns.

Or you can output all rows into the console, simply by doing:

`ali`

We wouldn't recommend this last option as it's not a particularly useful one, given that it floods your console, making it hard to see anything of use.

### Question Set One

Question 1: How many variables do we have in this dataset?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# There are two
# There are lots of ways to check this, you can do:
summary(ali)
ncol(ali) # how many columns (i.e. variables) in the ali dataset?
str(ali)  # what is the structure of ali?
head(ali) # will just show you
View(ali) # will also just show you
```

Question 2: What are the variable names?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# bite and length
# Above functions will give you this except for ncol()
```

Question 3: Are the names captilised?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# bite and length are both lower case
# A general tip - when you're creating your datasets, keep the variable/column
# names simple. You're going to be writing them out lots, so make typing easier
# for yourself by using short names that are all lower case.
# Definitely avoid names like "Alligator_Bite_Strength_PSI"
```

These questions may seem trivial, but honestly, they're the first questions we ask ourselves when we load in a dataset to `R` for the first time.

## Getting to know the data

Here's what you know about the data. The data contains `nrow(ali)` observations (sometimes refered to as samples) of `nrow(ali)` alligator bite strength, measured in pounds per square inch. All `nrow(ali)` alligators were measured in Parker Island Gator Farm, Florida, USA. For each bite strength, the length of the alligator, from nose to tail, was measured and is provided in centimeters (to one decimal place).

Before we move ahead with the data, let's take a second to become familiar with alligator bite strength. Using your search engine of preference, do a quick search to find out the bite strength of alligators. Specifically

When we do exploratory data analysis, we're specifically looking for things that look *weird*. In order to know if something is *weird*, we need to know what "normal" is; hence why it's a good idea to have some sort of notion of what a "normal" alligator's bite strength may be.

Write down the average bite strength of an alligator, measured in pounds per square inch and the average length of an alligator in centimeters.

With that done, let's move onto understanding our data. For starters, let's check if the data is sensibly structured. In `R` this means that numbers are numbers, words are characters, and so on. Basically, we want to make sure there weren't any bite measurements like `2174.94BIGBITE`, which will mess us up by meaning that `R` thinks that all `bite` values are characters and not numbers.

To do this quickly, we can use the `str()` function (short for "structure"):

```{r}
str(ali)
```
We see that `ali` is a `data.frame` (great - it should be!), with 78 observations (or samples) and two variables. The `$` is used to extract specific variables of a dataframe, so `ali$bite` will give you just the `bite` variable, hence why the second and third line are proceeded by `$`. From this, we can see that both the `length` and `bite` variables are being treated as numbers (the `num` part), and we then get a handful of example values for both `length` and `bite`.

So far, we're not seeing anything unusual which is great - it means we have not identified any problems that we need to fix.

If we so wanted to, we could extract the `length` variable from the dataset, by doing:

```{r}
ali$length
```

How would you extract the `bite` variable?

Let's use these extracted variables to calculate some summary statistics, starting with the mean (i.e. average). To do so, we use the `mean()` function. To find the average length of our alligators, we do:

```{r}
mean(ali$length)
```

We can see that the average crocodile in Parker Island Gator Farm is `r round(mean(ali$length), digits = )` cm long. 

Compare this value with the mean length of alligators you found from a quick internet search. A little bit longer than we might have expected but maybe it's something to do with these alligators coming from a farm, as opposed to wild ones. Maybe they're given lots of food to help them grow? Who knows, but we wouldn't expect our numbers to match exactly with whatever number we find online, right?

Do the same thing for `bite`; Calculate the average bite strength of our alligators, and let's see if there's anything weird going on there.

```{r}
mean(ali$bite)
```

Assuming you also found that alligator bite at roughly 2,000 pounds per square inch, our alligators seem to be biting *on average* with an extra 1,500 pounds per square inch. That seems like *a lot* more. 

This is something that should make us sit up and start paying a little more attention. We don't know there is a problem, but we have reason to believe there *might* be one. It's now up to us to prove that there *isn't* a problem.

There are lots of descriptive techniques we can use to check for odd things in our data, but a convenient "short cut" option is to use the `summary()` function. This function will provide a generic summary of our entire dataset if we supply `ali` as the sole argument. Do so now and let's see if we can track down what's going on in this data.

```{r}
summary(ali)
```

Keep in mind, with EDA where looking for anything *weird* in the data.

### Question Set Two

Question 1. How small is the smallest alligator in the farm for which we have data, and what's the biggest?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# 153.5 cm to 2293.0 cm
# Multiple ways to check this
range(ali$length)   # will give you the minimum and maximum
min(ali$length)     # will give you the minimum
max(ali$length)     # will give you the maximum
summary(ali)        # will give you this for all data in ali
summary(ali$length) # will give you summary of just length
```

Question 2. What are the minimum and maximum values for bite strength?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# 832.9 PSI to 123632.0 PSI
```

Question 3. Do any of these minimums or maximums seem strange?

```{r, include=answer_sheet, class.source="bg-success", eval = F}
# Yes! An alligator that's longer than a bus?!
# The same alligator, or a different one, that bites with
# 123,632 pounds per square inch, where the average is apparently 
# about 2,000?!
```

## Data detective

Hopefully you've identified that there is at least one questionable value in our dataset. What do we do now? Throw out the entire dataset? (No... The answer is "No you do not throw out the dataset" - at least not yet...). Let's try and figure out what's going on here, starting with finding the problem observations.

How do we do that?

Let's start with considering what we think a weird observation may look like. We can use the `summary(ali)` output to help us here. Starting with `length`, we know we have alligator that was apparently `r max(ali$length)` cm long (`r round(max(ali$length)/100, digits = 0)` m - that's longer than a bus!). How do we find that value in our dataset to see if there was anything else weird about it.

This is where we can use the square brackets `[]`. In `R`, square brackets allow us to extract specific observations from, in this case, a dataframe (technically, called "indexing"). To use them, first think how you would do this by hand:

1. We'd start by looking at the `ali` dataset
2. We'd then skim through the `ali` dataset, looking for any rows where alligators are `r max(ali$length)` cm long
3. Once we'd found these rows, then we'd look at all of the information related to those alligator

How does that translate into `R`? Almost exactly - we just write those three steps using code.

1. We'd start by looking at the `ali` dataset

```{r}
head(ali, n = 10) # just showing first 10 rows to avoid spam
```


2. We'd then skim through the `ali` dataset, looking for any rows where alligators are `r max(ali$length)` cm long

This is where we can make use of the square brackets as well as conditional statements. Conditional statements are where we can say "when this condition is met, do *something*". In our case, our condition is an alligator that's `r max(ali$length)` cm long. In `R` we can write this condition as (note that we use double equal signs to say "equal to"  in `R`):

```{r}
ali$length == 123632.0
```

The output is a list of `r nrow(ali)` values of `TRUE` and `FALSE` values, one for each row of our `ali` dataset. Any `TRUE` values indicate an alligator that is equal to `r max(ali$length)` cm long, and `FALSE` values are when that is not the case (in `R` we write "not equal to" as `!=`).

3. Once we'd found these rows, then we'd look at all of the information related to those alligator

Here is where we can use the square brackets. Square brackets allow us to extract specific rows, columns or just individual values from datasets. When using square brackets, keep in mind that the row number goes before the comma and column number goes after (e.g. `ali[row,column]`). If we don't specify a value for either `row` or `column`, `R` will give us all of the rows or all of the columns. For example, if we wanted to see all the information (i.e. all columns) for the 30th row, we would do:

```{r}
ali[30,]
```

We can use these square brackets, in combination with the conditional statement, to extract the entire row for the alligator that's `r max(ali$length)` cm long. What happens in this case is the `TRUE` and `FALSE` values act as a way to tell `R` which rows we want to see. If it's `TRUE`, then we want to see it. If it's `FALSE`, then don't show us. We can do this by writing:

```{r}
ali[ali$length == 2293,]
```

What do you make of this observation? We're expecting `length` to be about 200 to 400 centimeters (2 to 4 m), but this alligator is apparently 10 times bigger than that. And `bite` is about 100 times bigger. Neither seem to have a decimal point (or at least `R` has given them values of zero after the decimal point).

Why might that be the case? We'll let's see if we can find a "normal" observation to compare this giant alligator to. Let's assume the first observation is a normal one. Use the square brackets to extract the first row of data from the `ali` dataset.

```{r}
ali[1,]
```

Well, they're clearly different. The first alligator in our dataset looks pretty normal. Maybe a bit on the big side, but not weirdly so. There are only two possible explanations for this. Either there are giant alligators among us. Or, there has been a data entry mistake.

### Question Set Three

Question 1. What may have occurred to have caused this apparent data entry mistake?

```{r, include=answer_sheet, class.source="bg-success"}
# This "giant" alligator is the result of a data entry mistake whereby 
# the decimal points were missing for bite and length. The values should be 
# 229.3 cm long and	1236.32	pounds per square inch. If you look at the raw data, 
# you'll see it has no decimal points.
```

## A more refined detective

The way we identified the above data entry mistake is really useful, but what if we wanted to be a slightly more refined programmer in checking that observation? Basically, what if we didn't want to write out `r max(ali$length)` every single time?

What we can do instead, is use a function that returns the maximum alligator length. The function to do just that is called `max()`:

```{r}
max(ali$length)
```

We can use that when extracting the problematic variable by doing:

```{r}
ali[ali$length == max(ali$length),]
```

Which you can read as saying: "Within the `ali` dataset, extract all columns ("all columns" because we have not included anything after the comma in the square brackets) for any rows where the length of an alligator is equal to the maximum length of any alligator".

We can also adapt the conditional statement. For instance, if we wanted to extract all rows with alligators that were *greater* than the average, we could do:

```{r}
ali[ali$length > mean(ali$length),]
```

We now get data from more than one alligator as, unsurprisingly, more than one alligator meets our condition of being longer than average. We can also see the "data entry mistake alligator" using this method.

Let's keep refining this method by using a different function; `quantile()`. Go back and rerun the `summary(ali)` code and have a look at the quantiles (`1st Qu.` and `3rd Qu.`) that it reports. These represent the values above and below which contain 25% and 75% of the data. We can get these values ourselves by doing:

```{r}
quantile(ali$length, probs = c(0.25,0.75))
```

The `probs` argument is where we can specify what probabilities we would like quantiles for. In the above example, we've asked for the 25% (`0.25`) and 75% (`0.75`) quantiles. But because we've asked for two values, we need to "wrap" these in another function - the `c()` function. `c()` combines, or concatenates, more than one value together, hence how we were able to get to values of our the `quantile()` function. If we only wanted a specific quantile, we could write `quantile(ali$length, probs = 0.75)`. Because we only want one value, we don't need to use `c()`.

### Question Set Four

Question 1. Using `quantile()` extract all rows in the `ali` dataset where `bite` values are above the 3rd quantile.

```{r, include=answer_sheet, class.source="bg-success"}
ali[ali$bite > quantile(ali$bite, probs = 0.75),]
```

Question 2. Repeat the above, but for `bite` values *less than or equal to* the 33% quantile. If you're unsure how to specify *less than or equal to* have a quick search online or use your preferred LLM.

```{r, include=answer_sheet, class.source="bg-success"}
ali[ali$bite <= quantile(ali$bite, probs = 0.33),]
```

Question 3. Using the tools from above, can you identify any other possible data entry mistakes caused by missing decimal points?

```{r, include=answer_sheet, class.source="bg-success"}
# None of the other observations appear to have this "missing decimal point" error
```

## Correcting data

So. We've identified a problem. What now? Do we delete the observation? Declare that [*Sarcosuchus*](https://en.wikipedia.org/wiki/Sarcosuchus) are not extinct and are alive and well in Parker Island Gator Farm, Florida, USA (and have actually doubled in size)? Set up a facebook group to share our concrete and irrefutable evidence of *Sarcosuchus* and start producing a [film](https://www.facebook.com/cryptofilm/) to share this wonderful news with the world? Obviously not. That'd be silly.

But we do need to do something. Row 20 of our dataset is clearly a mistake but correcting it is a really sensitive topic. Part of the reason for that sensitivity is because a non-trivial amount of scientists have been found to actually *fake* their data (e.g. see [here](https://www.science.org/content/article/university-investigation-found-prominent-spider-biologist-fabricated-falsified-data), [here](https://www.science.org/content/article/does-ocean-acidification-alter-fish-behavior-fraud-allegations-create-sea-doubt), and most ironically [here](https://behavioralscientist.org/harvard-professor-under-scrutiny-for-alleged-data-fraud/) for examples).








