---
title: "Workshop 1 - Study Notes"
author: "Michelle Pinard & Deon Roos"
date: 'Last updated: `r Sys.Date()`'
output:
  word_document:
    toc: yes
    reference_docx: "../sbstemplate.docx"
---

# Purpose

Practice a repeatable EDA and data-cleaning workflow before any modelling. You loaded data, checked structures and ranges, spotted anomalies, and made transparent corrections in code.

# Project setup and reading data

Work inside RStudio, write code in a script, and run with `Ctrl/Cmd+Enter`. Set your working directory to whereever you saved your data using `setwd()`. Read in plain text data with headers using `read.table`.

Common load failures are misspelled paths, wrong file format, or a missing closing parenthesis.

```{r, eval = F}
setwd("C:/BI3010/Workshop1")
ali <- read.table("data/alligator_bite.txt", header = TRUE)
```

# First look

Check structure, variable types, and a few rows; then summarise the whole frame. Use `$` to pull a single column.

```{r, eval = F}
str(ali)
head(ali)
summary(ali)
ali$length
```

# Indexing essentials

Square brackets select rows, then columns: `my_data[rows, cols]`. Logical vectors filter rows; leaving `cols` blank returns all columns. Use `==`, `!=`, `<`, `<=`, `>`, `>=` for conditions.

```{r, eval = F}
ali[ali$length == 2293, ]
ali[1, ]
```

# Descriptive summaries you will use repeatedly

Range, min, max, mean, quantiles. Remember to use `na.rm` when NAs are present. Quantile probabilities are on [0, 1] (i.e. are between 0 and 1).

```{r, eval = F}
range(ali$bite)
min(ali$bite)
max(ali$bite)
mean(ali$bite)
median(ali$bite, na.rm = TRUE)
quantile(ali$bite, probs = c(0.25, 0.75))
```

# Finding extremes safely

Combine square brackets, conditions and functions to find values that meet your conditions.

```{r, eval = F}
ali[ali$length > mean(ali$length), ]
ali[ali$length <= 0, ]
```

# Typical data issues and transparent fixes

Check for differences in spelling and capitalisation, sentinel codes to NA (e.g. `999` to `NA`), sign errors (e.g. missing `-`), and missing decimals. Always show the rows you will change, perform the change, then re-summarise to make sure it worked. If it didn't work, quickly re-load the data to undo any mistakes.

```{r, eval = F}
# Spelling
chupa$city[chupa$city == "amarillo"] <- "Amarillo"

# Sentinel code
chupa$ufo[chupa$ufo == 9999] <- NA

# Sign error
chupa$alc[chupa$alc == -5] <- 5

# missing decimals (row known from EDA)
ali$length[20] <- ali$length[20] / 10
ali$bite[20]   <- ali$bite[20]   / 100

summary(ali)
summary(chupa)
```

# Strings and factors

Use `unique` to identify all unique levels of a factor.

```{r, eval = F}
unique(fast$species)
```
